flowCore a Bioconductor package for high throughput flow cytometry

### 2008-12-17 paper/bmc_article.tex
more changes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35949 bc3139a8-67e5-0310-9ffc-ced21a209358
add elements in the papaer, especially in the Results and Discussion section
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@36080 bc3139a8-67e5-0310-9ffc-ced21a209358

  \paragraph*{Results:} We developed a set of flexible open source
  computational tools (in the R package \Rpackage{flowCore}) to
  facilitate the analysis of these complex data. A key component of
  which is having suitable data structures that support the
  application of similar operations to a collection of samples or a
  clinical cohort. In addition, our software constitutes a shared and
  extensible research platform that enables colaboration between
  bioinformaticians, computer scientists, statisticians, biologists
  and clinicians. This platform will foster the developement of novel
  analytic methods for flow cytometry.

  \paragraph*{Conclusions:} 
  The software has been applied in the analysis of various data sets
  and its data structures have been proven to be highly efficient in
  capturing and organizing the analytic work flow. A number of
  additional Bioconductor packages successfully build on the
  infrastructure provided by \Rpackage{flowCore}, opening new avenues
  for flow data analysis.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry high content screening (FC-HCS) to
generate large, complex data sets in both basic and clinical research
applications \cite{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies \cite{mahnke2007omi}. In particular, the data often
need to be organized into groups of samples based on combinations of
additional covariates and similar operations need to be applied to
these groups in a transparent and reproducible manner. Furthermore,
the growing depth of knowledge in the field of immunology, for
instance the charactarization of distinct human T-cell subpopulation
\cite{Appay}, clearly argues for more systematic approaches.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

@@

Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. Most of the currently available
analysis tools are designed to reflect this work flow.  We believe
that these approaches, in addition to being expensive and labor
intensive, do not fully address the highly complex nature of FCM data;
in particular, they disregard many of the fundamental aspects of the
data, such as sample groups or cohorts, the underlying distribution or
its high-dimensional nature. Furthermore, the subjective character of
manual analyses are a major obstacle to reproducibility. In a recent
study of flow cytometric standardization involving 15 institutions,
the mean inter-laboratory coefficient of variation ranged from 17 to
44\%, even though preparation was standardized and performed using the
same samples and reagents at each site \cite{Maecker2005}. For FC-HCS
data, unassisted manual inspection is extremely time consuming, and
robust statistical methods need to be developed to point investigators
to interesting aspects of the data, or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that collaboration with
other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

@@

%% Perhaps in Conclusion
%% Consequently, we are optimistic that this
%% software platform will play a critical role in addressing the
%% bottlenecks described above that limit advances in the application of
%% FC-HCS.
%%

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
The computational tools we have developed are distributed in the R
software language \cite{Rmain} as the Bioconductor \cite{BIOC} package
\Rpackage{flowCore}.  The package \Rpackage{flowCore} is a freely
available, highly functional, and extensible FCM data analysis
platform that enables researchers to efficiently handle FC-HCS data
and encourages open development of tools for their coherent
analysis. In our implementation of \Rpackage{flowCore} we rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second is the importance of a modular
architecture that allows for many developers to extend and use the
underlying infrastructure and to combine tools in complex work flows.
\Rpackage{flowCore} implements such computationally efficient data
structures and a range of specialized methods addressing all
components of a typical FCM analysis work flow, including
compensation, transformation, and gating. \Rpackage{flowCore} runs on
Windows, Mac OS X, and Linux/Unix operating systems. 
%% In this paper, we describe the functionality and methods underlying
%% \Rpackage{flowCore}, and provide examples of their use.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes them impractical to integrate
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (e.g., Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

@@

\Rpackage{flowCore} is not a GUI-driven software, and all operations
are done using a command line interface.  It is possible to add a more
elaborate user interface on top of this infrastructure, however the
focus in this paper is on a programmatic approach to enable the
convenient development of novel analysis methods and automation of
complex analysis approaches.  By taking the burden of data management
from the programmer, and by providing well-defined programming
interfaces (APIs), it is possible to readily test new ideas and to
easily extend the framework's functionality.

\subsection*{Basic Data Structures}
\subsubsection*{flowFrame}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other successful Bioconductor
packages. All information from a single FCS file, i.e., the collection
of events and the accompanying metadata, is stored in one single
container. We call the structure that hold this data a
\Rclass{flowFrame}. Raw data values as well as associated metadata of
a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
the metadata consist of descriptors of the stains used in the
experiment and the respective measurement channels, information about
compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. During
the creation of a \Rclass{flowFrame}, a number of quality checks are
performed to ensure data integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple samples, hence we introduce the
concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a microtitre plate.  \Rclass{flowSets} manage the consistent
application of operations on the individual \Rclass{flowFrames} and
shift the burden of keeping score of the metadata from the user to the
infrastructure, thus reducing the risk of errors (\textit{e.g.},
mixups of sample labels). Crucial operations like taking subsets, data
transformations and gating or computation of summary statistics are
greatly facilitated and all relevant annotation information is
constantly passed on along the analysis pipeline.

The \Rclass{flowSet} structure can be readily extended to incorporate
the potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials, where hundreds of patients might
provide samples at different time points over the course of the
experiment. The \Rclass{flowSet} data structure is one of the key
features in the \Rpackage{flowCore} package and it is fundamental to
the implementation of many of the high level functionalities such as
quality assessment and control, visualization and automateds gating.

\subsection*{Standard flow operations}
Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

\subsubsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \cite{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantial or arcsinh,
see Table~\ref{table1} for a complete list). Furthermore, the design
of the R language makes it easy to define arbitrary functions to apply
to the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate single stain controls.

@@

The ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, for example the proportions
of events falling in a single gate or in a combination of
gates. Complex combinations and hierarchies of gates can be captured
in multi-step gating strategies. The definition of gates in
\Rpackage{flowCore} follows the Gating-ML CR, thus any
\Rpackage{flowCore} gating strategy can be reproduced by any other
software that also adheres to the standard and \textit{vice versa}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
\subsection*{From quality assessment to batch gating}
The \Rpackage{flowCore} package has been successfully applied in the
analysis of several data sets, both originating from clinical trials
\cite{brinkman2007hcf} and drug discovery experiments
\cite{gasparetto2004ice}. A complete description of the
\Rpackage{flowCore} software is beyond the scope of this publication.
Much more comprehensive documentation and users guide information with
programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the \Rpackage{flowQ}
package to create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
\Rclass{flowSet} object ``dat'', the following simple line of code
produces the output shown in Figure~2\ref{flowQ}:

@@

Static gating for all samples in a high-throughput FCM experiment is
often impossible, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating has the potential to estimate
the gating regions from the underlying data, thus providing a fast
objective solution to the analysis of potentially very large and
diverse data sets \cite{lo2008agf}. One of the automated gating
methods implemented in \Rpackage{flowCore} is based on identifying
areas of significant curvature in a kernel density estimate of the
data \cite{wand2008}. Assuming that the regions of interest are of
high density, the software is able to reliably detect them in a one or
two-dimensional density landscape.

@@

Kernel density estimation is a well-known problem in statistical
computing, and a lot of effort has been invested in the development of
good software to address it. The modular design of \Rpackage{flowCore}
allows to easily integrate these existing solutions into our
framework. In this example, we directly use R code from the
\Rpackage{feature} package \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in \Rpackage{flowCore} is available to other package
authors.

We can chose one of the many visualization option from the
\Rpackage{flowViz} package to plot the results of the recent filtering
operation. A very basic matrix of density plots is shown in Figure~3
\ref{xyplot}, where each panel in the matrix represents the data of
one individual patient.

@@

\subsection*{Related flow packages}
In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \cite{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \cite{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitating the
handling of spatial metadata, is under development.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
Through \Rpackage{flowCore}, we have provided the FCM community with
an open source, freely available, highly functional, standards
compliant development and analysis platform for high throughput data
analysis.  We hope to foster collaborative development of new analysis
methods and to facilitate the transition of these new methods to the
larger flow community.  Our experience has been that such
collaborative effort has proven beneficial for a number of different
biological and computational biology challenges, greatly elevating
their applicability.  We hope that our framework will be the
foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent further development and deployment of FC-HCS to
increasingly complex and important scientific and clinical
applications.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
Project name: flowCore
Project home page: http://bioconductor.org 
Operating system(s): The flowCore package can be installed on all the 
platforms supporting R. These include a wide variety of UNIX platforms, Windows 
and MacOS.  
Programming language: R 
Licence: The Artistic License, Version 2.0

@@

  \subsection*{Figure 2 - Quality assessment}
    \label{flowQ}%
        HTML quality assessment report generated by the flowQ package for a
        subset of data from an experiment focussing on Graft-Versus-Host
        Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
        the set, columns to different quality checks.

    \subsection*{Figure 3 - Quality assessment}
    \label{xyplot}{
        Scatterplot matrix of a single \Rclass{flowSet} from an experiement
        focussing on immune tolerance following kidney
        transplantation. Outlines of the gating regions identified by an
        automated gating operation are added on top of the density
        representation of the data.}

@@

  \subsection*{Table 2 - Filter and gate classes}
  \begin{table}[ht]
    \caption{\label{table2} Filter and gate classes implemented in
      \Rpackage{flowCore}.} 
    \begin{center}
      \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|c|}{Gates} \\
        \hline
        rectangleGate & n-dimensional rectangular regions \\
        quadGate & quadrant regions in two dimensions \\
        polygonGate & polygonal regions in two dimensions \\
        polytopeGate & generalization of polygon in n dimensions \\
        ellipsoidGate & n-dimensional ellipsoid region \\
        \hline
        \multicolumn{2}{|c|}{Filters} \\
        \hline
        sampleFilter & random sub-sampling\\
        expressionFilter & results of a boolean expression \\
        kmeansFilter & K-means clustering \\
        norm2Filter & bivariate normal distribution \\
        curv1Filter & local density regions in 1D \\
        curv2Filter & density regions in 2D \\
        timeFilter & abnormal data acquisition over time \\
        \hline
        filterSet & gating strategies \\
        \hline
      \end{tabular}
    \end{center}
  \end{table}

>>>>

  \paragraph*{Results:} We developed a set of flexible open source
  computational tools (in the R package \Rpackage{flowCore}) to
  facilitate the analysis of these complex data. A key component of
  which is having suitable data structures that support the
  application of similar operations to a collection of samples or a
  clinical cohort. In addition, our software constitutes a shared and
  extensible research platform that enables collaboration between
  bioinformaticians, computer scientists, statisticians, biologists
  and clinicians. This platform will foster the development of novel
  analytic methods for flow cytometry.

  \paragraph*{Conclusions:} 
  The software has been applied in the analysis of various data sets
  and its data structures have proven to be highly efficient in
  capturing and organizing the analytic work flow. Finally, a number
  of additional Bioconductor packages successfully build on the
  infrastructure provided by \Rpackage{flowCore}, open new avenues for
  flow data analysis.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry (FCM) to generate large, complex
data sets in both basic and clinical research applications
\cite{brinkman2007hcf}. A serious bottleneck in the interpretation of
existing studies and the application of flow cytometry high content
screening (FC-HCS) to even larger, more complex problems is that data
management and data analysis methods have not advanced sufficiently
far from the methods developed for applications of FCM
 to small-scale,
tube-based studies \cite{mahnke2007omi}. In particular, the data often
need to be organized into groups of samples based on combinations of
additional covariates and similar operations need to be applied to
these groups in a transparent and reproducible manner. Furthermore,
the growing depth of knowledge in the field of immunology, for
instance the characterization of distinct human T-cell sub-population
\cite{Appay}, clearly argues for more systematic approaches.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

@@

Traditionally, for the majority, FCM experiments were being analyzed
by manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. Most of the currently available
analysis tools are designed to reflect this work flow.  We believe
that these approaches, in addition to being expensive and labor
intensive, do not fully address the highly complex nature of FCM data;
in particular, they disregard many of the fundamental aspects of the
data, such as sample groups or cohorts, the underlying distribution or
its high-dimensional nature. Furthermore, the subjective character of
manual analyses are a major obstacle to reproducibility. In a recent
study of flow cytometric standardization involving 15 institutions,
the mean inter-laboratory coefficient of variation ranged from 17 to
44\%, even though preparation was standardized and performed using the
same samples and reagents at each site \cite{Maecker2005}. For FC-HCS
data, unassisted manual inspection is extremely time consuming, and
robust statistical methods need to be developed to point investigators
to interesting aspects of the data, or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that collaboration with
other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
The computational tools we have developed are distributed in the R
software language \cite{Rmain} as the Bioconductor \cite{BIOC} package
\Rpackage{flowCore}.  The package \Rpackage{flowCore} is a freely
available, highly functional, and extensible FCM data analysis
platform that enables researchers to efficiently handle FC-HCS data
and encourages open development of tools for their coherent
analysis. In our implementation of \Rpackage{flowCore} we rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second is the importance of a modular
architecture that allows for many developers to extend and use the
underlying infrastructure and to combine tools in complex work flows.
\Rpackage{flowCore} implements such computationally efficient data
structures and a range of specialized methods addressing all
components of a typical FCM analysis work flow, including
compensation, transformation, and gating. \Rpackage{flowCore} runs on
Windows, Mac OS X, and Linux/Unix operating systems.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures.
In addition, functions or methods that work on those classes allow for
interaction and manipulation. Many of the currently available software
solutions offer only limited support for such self-contained
structures, or make use of containers like XML or binary storage
formats that are designed specifically for particular user interfaces
and hence not easily amenable to programmatical access. In addition,
the closed-source nature of these products often makes them
impractical to integrate into analysis pipelines. In this manuscript
we describe classes for FCM data analysis and their implementation in
R, however, they could just as easily be implemented in any other
language (e.g., Java, C++). Software written in those languages could
use similar data structures, thereby simplifying communication and the
interchange of data between analysis tools.

%% I think switching the last 2 paragraphs make more sense as the 2nd
%% one of this section talk about other closed-source soft etc.
%% But up to you
\Rpackage{flowCore} does not provide a graphical user interface
(\textit{i.e.}, is not a GUI-driven software) and all operations are
done using a command line interface.  It is possible to add a more
elaborate user interface on top of this infrastructure, however the
focus in this paper is on a programmatic approach to enable the
convenient development of novel analysis methods and automation of
complex analysis approaches.  By taking the burden of data management
from the programmer, and by providing well-defined application
programming interfaces (APIs), it is possible to readily test new
ideas and to easily extend the framework's functionality.

@@

\subsection*{Basic Data Structures}
\subsubsection*{flowFrame: sample unit}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other successful Bioconductor
packages. All information from a single FCS file, \textit{i.e.}, the
collection of events and the accompanying metadata, is stored in one
single container. We call the structure that hold this data a
\Rclass{flowFrame} (Figure~1\ref{fig1:FrameWork}). Raw data values as
well as associated metadata of a \Rclass{flowFrame} can be accessed
programmatically. Most commonly, the metadata consist of descriptors
of the stains used in the experiment and the respective measurement
channels, information about compensation performed at the instrument
side and any additional keywords the user deems to be important to
annotate the data. During the creation of a \Rclass{flowFrame}, a
number of quality checks are performed to ensure data integrity.

\subsubsection*{flowSet: a collection of  \Rclass{flowFrames}}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple samples, hence we introduce the
concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet} (Figure~1\ref{fig1:FrameWork}). A \Rclass{flowSet} is
a container for multiple \Rclass{flowFrames} along with relevant
information associated with each individual frame such as descriptions
of the cell sample, the treatment to which the sample was subjected,
or the location of that sample in a microtitre plate.  The objects are
self-contained and can be shipped to other computers, platform
independently. \Rclass{flowSets} manage the consistent application of
operations on the individual \Rclass{flowFrames} and shift the burden
of keeping score of the metadata from the user to the infrastructure,
thus reducing the risk of errors (\textit{e.g.}, mixups of sample
labels). Crucial operations like taking subsets, data transformations
and gating, or computation of summary statistics are greatly
facilitated and all relevant annotation information is constantly
passed on along the analysis pipeline.

The \Rclass{flowSet} structure can be readily extended to incorporate
the potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials, where hundreds of patients might
provide samples at different time points over the course of the
experiment. The \Rclass{flowSet} data structure is one of the key
features in the \Rpackage{flowCore} package and it is fundamental to
the implementation of many of the high level functionalities such as
quality assessment and control, visualization and automated gating.

\subsection*{Standard flow operations}
Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, most often in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. Typically, the results of these operations are
again \Rclass{flowFrame} or \Rclass{flowSet} objects. While
transformation, and to a certain extent compensation, are fairly
routine operations with only limited potential for improvement, being
able to implement new methodologies for gating of FCM data, and
extend the capabilities of \Rpackage{flowCore} through object
oriented programming are features that clearly sets our framework
apart from other FCM analysis tools. By factoring out as much of the
bookkeeping as possible, programmers can focus on the actual
operations rather then having to deal with the tedious details of data
integration and access. Third-party methods can act on their own as
first-class citizens in the analysis framework, without breaking the
work flow or the basic infrastructure. This design allows for the
straightforward extension of \Rpackage{flowCore}'s capabilities, and
has already fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

\subsubsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \cite{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponential or arcsinh,
see Table~\ref{table1} for a complete list). Furthermore, the design
of the R language makes it easy to define arbitrary functions to apply
to the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation, that corrects for
fluorescence spillover originating from the inherent overlap of
emission spectra from antibody fluorescent labels, is available for
both \Rclass{flowFrames} and \Rclass{flowSets}. In addition, the
software offers functionality to compute spillover or compensation
matrices from a set of appropriate single stain controls.

@@

The ability to programmatically access gates is a prerequisite for
semi-automated or automated gating. By utilizing an unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, as for instance the
proportions of events falling in a single gate or in a combination of
gates. Complex combinations and hierarchies of gates can be captured
in multi-step gating strategies. The definition of gates in
\Rpackage{flowCore} follows the Gating-ML CR, thus any
\Rpackage{flowCore} gating strategy can be reproduced by any other
software that also adheres to the standard and \textit{vice versa}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
\subsection*{From quality assessment to batch gating}
The \Rpackage{flowCore} package has been successfully applied in the
analysis of several data sets, both originating from clinical trials
\cite{brinkman2007hcf} and drug discovery experiments
\cite{gasparetto2004ice}. A complete description of the methods
implemented in \Rpackage{flowCore} is beyond the scope of this
publication.  Much more comprehensive documentation and users guide
information with programmatic examples are available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html}),
as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

Data analysis for most experiments usually begins with a quality
assurance step. In a FCS analysis work flow, we can use functionality
from the \Rpackage{flowQ} package, build upon \Rpackage{flowCore}, to
create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
\Rclass{flowSet} object ``dat'', using for instance the
\Rclass{read.flowSet} method, the following simple lines of code
produce the output shown in Figure~2\ref{flowQ}:

@@

According quality of the measurements, the next steps of a FCS
analysis work flow are potentially the compensation and transformation
of the data. Once again \Rpackage{flowCore}'s data structure and its
methods allow a flexible processing. One can use the basic
transformation and compensation functions implemented in
\Rpackage{flowCore} (Table~1) or develop its own approaches that could
then be apply to \Rclass{flowFrame} or \Rclass{flowSet}.

Ultimately, a flow cytometry experiment aims at identifying and
characterizing cell population of biological interest, using static
gating or data-driven procedures (Table~2). Static gating for all
samples in a high-throughput FCM experiment is often impossible, since
the measured variables tend to vary between different treatments, over
time or between different experiment batches. Automated or data-driven
gating has the potential to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
\Rpackage{flowCore} is based on identifying areas of significant
curvature in a kernel density estimate of the data
\cite{wand2008}. Assuming that the regions of interest are of high
density, the software is able to reliably detect them in a one- or
two-dimensional density landscape.

@@

Kernel density estimation is a well-known problem in statistical
computing, and a lot of effort has been invested in the development of
good software to address it. The modular design of \Rpackage{flowCore}
allows to easily integrate these existing solutions into our
framework. In this example, we directly use R code from the
\Rpackage{feature} package \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionalities
implemented in \Rpackage{flowCore} are available to other package
authors.

Finally, we can chose one of the many visualization options from the
\Rpackage{flowViz} package to plot the results of the recent filtering
operation. A very basic matrix of density plots is shown in Figure~3
\ref{xyplot}, where each panel in the matrix represents the
fluorescent measurements of two channels for one individual patient.

@@

\subsection*{Related flow packages}
In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. As exemplified in the previous section, the
\Rpackage{flowViz} package \cite{sarkar2008ufv} provides sophisticated
data visualization tools, that make use off multivariate trellis
plotting \cite{lattice}.  These functions can be used to quickly
generate customized plots for extended cytometry data sets for both
direct data inspection and quality control.  The objects metadata
information can be used to arrange the layout and composition of the
plots.  Furthermore, the design and the API of the visualization
software is very generic, and users can readily extend its
capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow cytometry
data analysis.

More recently, \cite{lo2008agf} have developed an automatic gating
approach via robust model-based clustering using \Rpackage{flowCore}'s
data model and infrastructure which is implemented in the Bioconductor
package \Rpackage{flowClust}. Another package, \Rpackage{plateCore},
providing more specialized support for experiments conducted on
microtitre plates and facilitating the handling of spatial metadata,
is under development.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
Through \Rpackage{flowCore}, we have provided the FCM community with
an open source, freely available, highly functional, and standards
compliant, development and analysis platform for high throughput data
analysis.  We hope to foster collaborative development of new analysis
methods and to facilitate the transition of these new methods to a
larger flow community.  Our experience has been that such
collaborative effort has proven beneficial for a number of different
biological and computational biology challenges, greatly elevating
their applicability.  We hope that our framework will be the
foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent further development and deployment of FC-HCS to
increasingly complex and important scientific and clinical
applications.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
Project name: flowCore ;
Project home page: http://bioconductor.org ;
Operating system(s): The flowCore package can be installed on all the 
platforms supporting R. These include a wide variety of UNIX platforms, Windows and MacOS.;  
Programming language: R ;
Licence: The Artistic License, Version 2.0 .

@@

  \subsection*{Figure 2 - Quality assessment}
    \label{flowQ}%
        HTML quality assessment report generated by the flowQ package for a
        subset of data from an experiment focusing on Graft-Versus-Host
        Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
        the set, columns to different quality checks.

    \subsection*{Figure 3 - Quality assessment}
    \label{xyplot}{
        Scatterplot matrix of a single \Rclass{flowSet} from an experiment
        focusing on immune tolerance following kidney
        transplantation. Outlines of the gating regions identified by an
        automated gating operation are added on top of the density
        representation of the data.}

@@

  \subsection*{Table 2 - Filter and gate classes}
  \begin{table}[ht]
    \caption{\label{table2} Filter and gate classes implemented in
      \Rpackage{flowCore}. Filters are automated, data driven procedures. Gates are static, user-defined methods.} 
    \begin{center}
      \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|c|}{Gates} \\
        \hline
        rectangleGate & n-dimensional rectangular regions \\
        quadGate & quadrant regions in two dimensions \\
        polygonGate & polygonal regions in two dimensions \\
        polytopeGate & generalization of polygon in n dimensions \\
        ellipsoidGate & n-dimensional ellipsoid region \\
        \hline
        \multicolumn{2}{|c|}{Filters} \\
        \hline
        sampleFilter & random sub-sampling\\
        expressionFilter & results of a boolean expression \\
        kmeansFilter & K-means clustering \\
        norm2Filter & bivariate normal distribution \\
        curv1Filter & local density regions in 1D \\
        curv2Filter & density regions in 2D \\
        timeFilter & abnormal data acquisition over time \\
        \hline
        filterSet & gating strategies \\
        \hline
      \end{tabular}
    \end{center}
  \end{table}
